---
title: "The case for a pipe assignment operator in R"
author: David Hugh-Jones
format: html
table-of-contents: true
---



# Assignment in R

Here are two key facts about R: 

* R is *pass by value*;
* R has complex, expressive idioms for assigning to subsets.

Pass by value means that you can almost never change an
object like this:

``` r
do_something(obj)
```

That just gives the function a copy of `obj`; the original
`obj` will be unchanged. Instead, you write:

``` r
obj <- do_something(obj)
```

Complex subsetting means that not just simple objects can be
on the left hand side of assigments. You can also write:

``` r
data[condition, cols] <- do_something(data[condition, cols])
```

This replaces the rows where `condition` is `TRUE` and the columns
specified in `cols`. And there are lvalue functions too:

``` r
names(data) <- str_remove(names(data), "\\d+$")
```

Combining these can lead to expressive, but complex, transformations:

``` r
names(data)[1:2] <- paste0(names(data)[1:2], "_suffix")
```

The duplication here makes the code hard to read, and easy to
break, for example by updating the subset index on the left
hand side and forgetting to do the same on the right. You
might prefer to break this down for legibility:

``` r
index <- 1:2
new_names <- names(data)[index]
new_names <- paste0(new_names, "_suffix")
names(data)[index] <- new_names
```

# The pipe operator

R 4.1.0 introduced the native pipe operator. A simple syntax
transformation converts

```r
x |> mean(na.rm = TRUE)
```

into

``` r
mean(x, na.rm = TRUE)
```

The `x` on the left enters the function on the right, as its first argument.

In terms of language capabilities, this was a no-op: R
didn't gain the ability to do anything it couldn't already do.
But by putting the subject of the code first, and operations 
done to it in order, the pipe can make R code much more expressive:

``` r
library(dplyr)
mtcars |> 
  group_by(gear) |> 
  mutate(kw = hp * 0.746) |> 
  summarize(mean_kw = mean(kw))
```

# Pipe assignment

I'd like to propose a new **pipe assignment** operator for R.
The pipe assignment operator would convert:

``` r
obj <|> do_something(...)
```

into: 

``` r
obj <- do_something(obj, ...)
```

Equivalently, `obj <|> do_something()` can be read as
`obj <- obj |> do_something()`.

Having this operator would make code simpler and more
expressive, and reduce bugs. For example, the complex
assignments above could be rewritten:

```r
data[condition, cols] <- do_something(data[condition, cols])
# becomes
data[condition, cols] <|> do_something()

names(data)[1:2] <- paste0(names(data)[1:2], "_suffix")
# becomes
names(data)[1:2] <|> paste0("_suffix")
```

The new versions are easier to read: you only need to parse
one complex expression, instead of two. And they are harder to
break, because the left and right hand sides don't need to be
kept in sync.

# Some evidence

How useful would pipe assignment be? I used a 
[newly-created database of R code](https://github.com/hughjonesd/codesamples) to look for
assignments. The tables below show statistics from about
26,000 code snippets, from GitHub repositories,
StackOverflow questions and R package examples ([code](find-pipeable-assignments.R)).
I counted what proportion of 
assignments were *pipeable*, i.e. of the form
`a <- foo(a, ...)`. I also counted how many were *complex*,
i.e. like `a$blah <- foo(a$blah, ...)` or 
`a[subset] <- foo(a[subset], ...)`.

```{r}
#| file: find-pipeable-assignments.R
#| cache: true
#| echo: false
#| output: false
```

```{r}
#| echo: false
library(huxtable)
suppressPackageStartupMessages(library(dplyr))

summaries <- bind_rows(
  GitHub = gh_summary,
  StackOverflow = so_summary,
  Examples = ex_summary,
  .id = "Source")

summaries |> 
  rename(
    Source = Source,
    N = n,
    `Pipeable (%)` = prop_pipeable,
    `Pipeable and complex (%)` = prop_both
  ) |> 
  as_hux() |> 
  set_number_format(everywhere, 3:4, fmt_percent(1)) |> 
  set_caption("Statistics of code snippets")
```


About 4-10% of all assignments could be simplified using an
assignment pipe. That is a big potential gain, because
assignment is probably the single most common operation in R
code. The GitHub data contains about 9000
uses of `<-`, of which about 900 could be simplified. For
comparison, it contains about 740 uses of the function
`median`. Even in the R examples, which are meant to be
concise, 4% of assignments were of the form `x <- foo(x)`.

A smaller proportion of assignments are both pipeable and complex. Here are some standout cases of long, complex assignments:

``` r
# from StackOverflow questions:

w[(midpoint - formerBreak):(midpoint + formerBreak), 
  (midpoint - formerBreak):(midpoint + formerBreak)] <- 
  w[(midpoint - formerBreak):(midpoint + formerBreak), 
    (midpoint - formerBreak):(midpoint + formerBreak)] * 
    (wOld == 0)

paretoData$EffectNames = factor(paretoData$EffectNames, 
  levels = c("Pull Back(A)", "Hook(B)", "Peg(C)", "AB", "BC",
             "AC", "ABC"))
             
mean.vis.aud.long$Visual = Relevel(mean.vis.aud.long$Visual,
                ref = c("LeftCust", "SIMCust", "RightCust"))
                
data[data$category == category, ]$value = 
  data[data$category == category, ]$value /
  data[data$category == category & 
       data$year == baseYear,]$value[[1]]
       
wardmap@data = data.frame(
                 wardmap@data, 
                 mayor2015[
                   match(wardmap@data[, "WARD"], 
                   mayor2015[, "WARD"]), ])
```

Got that 😉? Note that the last expression could be rewritten
as

``` r
wardmap@data <|> (function (x) {
  data.frame(x, 
             mayor2015[
               match(x[, "WARD"], mayor2015[, "WARD"]), 
             ]))()
```

It could be done even more simply if the pipe assignment
operator had a way to pass the left hand side into multiple places. (And if there were a nice way to write anonymous function calls in R - but that's another topic.)

StackOverflow questioners might not be good at writing simple
code. What about Github? Here's some examples:

``` r
data_for_model$factor_parameter_dictionary = 
    bind_rows(data_for_model$factor_parameter_dictionary,    
        mutate(tibble(design_matrix_col = 
            names(select_if(distinct(select(.data_spread, 
               all_of(parse_formula(formula)))), function(x) 
                       is.numeric(x)))), 
                       factor = design_matrix_col))
                       
                       
HillTN95[(1 + nrow(BasinSF) * (h - 1)):(1 + nrow(BasinSF) * (h - 1) + (nrow(BasinSF) - 1)), ] = 
  HillTN95[(1 + nrow(BasinSF) * (h - 1)):(1 + nrow(BasinSF) *
    (h - 1) + (nrow(BasinSF) - 1)), ] # if you are lost, this is the end 
                                      # of the original left hand side...
    [order(HillTN95[(1 + nrow(BasinSF) * (h - 1)):(1 +
      nrow(BasinSF) * (h - 1) + (nrow(BasinSF) -
      1)),]$Replicate),] # ... and here it is again
```

I'm not saying duplicated variables are the *only* problems 
of this code. But duplicated variables certainly
are not helping. 

But even experienced R coders could benefit from this idiom.
Here's more GitHub examples, with some famous names:


``` r
# tidyverse/tibble
x[needs_ticks] <- tick(x[needs_ticks])

# RConsortium/S7
signature[sig_is_union] <- lapply(signature[sig_is_union], "[[", "classes")
signature[!sig_is_union] <- lapply(signature[!sig_is_union], list)

# RevolutionAnalytics/dplyr-spark
mask = sapply(values, is.factor)
values[mask] = lapply(values[mask], as.character)
mask = sapply(values, is.character)
values[mask] = lapply(values[mask], encodeString)

# r-forge/optimizer
expr[[i]] <- nlsSimplify(expr[[i]], simpEnv, verbose = verbose)

# tidyverse/ggplot2
colnames(coef) <- to_lower_ascii(colnames(coef))

# cran/Matrix
x@x <- zapsmall(x@x, digits = digits)

# wch/r-source
# Yes, this is the R source code itself!
x$sec <- trunc(x$sec)
...
res$ar <- aperm(res$ar, c(2L,3L,1L))
```

Lastly, here's some example code from R packages:

``` r
# jrSiCKLSNMF::CalculateUMAPSickleJr
SimSickleJrSmall<-CalculateUMAPSickleJr(SimSickleJrSmall,
umap.settings=umap.settings)
SimSickleJrSmall<-CalculateUMAPSickleJr(SimSickleJrSmall,
umap.settings=umap.settings,modality=1)
SimSickleJrSmall<-CalculateUMAPSickleJr(SimSickleJrSmall,
umap.settings=umap.settings,modality=2)


# rms::predictrms
dat <- cbind(dat, predict(fit, dat, se.fit=TRUE))


# oce::`[[<-,section-method`        
section[["sectionId"]] <- toupper(section[["sectionId"]])
...
section[["station", 10]][["temperature"]] <-
    1e-3 + section[["station", 10]][["temperature"]]


# inTrees:Num2Level
rfList$list[[i]][, "prediction"] <-   
  data.frame(dicretizeVector(rfList$list[[i]][, "prediction"],
                             splitV))
```

Remember, examples are meant to be short and clear!

In all of these cases, I think `<|>` could make the code clearer, less repetitive and more readable.


# Avoiding hacks

Another advantage of an assignment pipe is that it obviates
the need for hacks to get round it. One example, arguably, is
dplyr's `mutate()` function. Don't get me wrong, I love dplyr
and it really changed R data manipulation for the better. But
one of its advantages is just that:

``` r
... |> mutate(foo = foo + 1) |> ...
```

is easier to read than

``` r
data$foo <- data$foo + 1
```

And while mutate itself is powerful, it can only deal with
one column at a time. When you want to change multiple columns, you have to use `across`, which looks like this:

``` r
   ... |> 
       mutate(
         across(cols, ~ foo(.x)
       ) |> ....
```

I'm [not a fan](https://medium.com/p/ae364da6a46f) of `across`, I think it's complex to read
and understand (in fact, I got it wrong in the first version
of the code above). But the key point is that it is a
workaround for:

```r
df[cols] <|> foo()
```



# Implementing it

If it's such a good idea, why don't I submit a patch?

Simply, I can't code C, certainly not well enough to deal 
with the R source. 

**Update**: but in the spirit of laziness, impatience and
hubris, here is a [patch](pipe-assignment.patch) and
see [this commit on github](https://github.com/hughjonesd/r-source/commit/de9b934ae89e59685f611022344570e0b51f2b3b). Output:

``` r
a <- 1:5
a <|> mean()
a
## [1] 3

a <- 1:5
a[1:2] <|> rev()
a
## [1] 2 1 3 4 5

names(a) <- letters[1:5]
names(a)[3:4] <|> paste0("-suffix")
a
##       a        b c-suffix d-suffix        e 
##       2        1        3        4        5 

# not everything works:
a <|> sqrt() |> mean()
## Error in sqrt() : 0 arguments passed to 'sqrt' which requires 1
```


I *believe* that the assignment pipe could
be implemented relatively simply, like the original native
pipe `|>`, as a syntax transformation.

So, I'm hoping that R Core will consider this possibility.

Meanwhile, R users can always import the magrittr package's
`%<>%` operator, which does what I'm calling for:

```{r}
library(magrittr)
x <- 1:5
x %<>% mean()
x
```

The magrittr `%<>%` is not implemented at R source level,
which makes it relatively slow and complex. I think the
advantage of `<|>` is just the same as the advantage of `|>`
versus `%>%`: it can be native and fast.

If you have comments or thoughts, drop me a line. I'm
@davidhughjones on X, or davidhughjones on gmail.

Merry Christmas/Season's Greetings and a Happy New Year!