---
title: "Making RmlxStats fast and accurate"
author: "David Hugh-Jones"
date: "2026-05-20"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(mypaintr)
```

[mlx](https://mlx-framework.org) is a python/C/C++ API for GPUs on modern Macs.
[Rmlx](https://hughjonesd.github.io/Rmlx) is an R interface to mlx. The core
idea behind Rmlx is simple: statistics uses a lot of matrix operations, the GPU
can do matrix operations fast, so we should let R have access to the GPU.

[RmlxStats](https://hughjonesd.github.io/RmlxStats) is a showcase for
Rmlx: it rewrites some well-known R model fitting functions using mlx as a
backend. These include `mlxs_lm()`, `mlxs_glm()`, `mlxs_prcomp()` for
principal components analysis, and `mlxs_glmnet()` which is inspired by
`glmnet()` from the glmnet package. The aim of RmlxStats is to learn what Rmlx
can do, and to provide fast versions of basic statistics.

mlx has some nice advantages: in particular, you can switch between GPU and CPU
operations without moving data around, which is helpful for many statistics 
operations. Still, it isn't a perfect solution. Not all operations work on the
GPU yet: `solve()` is CPU-only.

Despite this, mlx can be pretty fast:


  ```r
  # On my machine
  > system.time(lm(arr_delay ~ dep_delay + factor(paste(month,day)), 
                   data = nycflights13::flights))
     user  system elapsed 
   31.769   0.544  32.764 

  > system.time({
      fit <- mlxs_lm(arr_delay ~ dep_delay + factor(paste(month,day)), 
                     data = nycflights13::flights)
      Rmlx::mlx_eval(fit$coefficients)
    })
     user  system elapsed 
    4.274   0.739   3.351 
  ```
 
That snippet shows off another feature of mlx: evaluation is lazy, so
work isn't done until you evaluate the final result.
 
# Fast but imprecise
 
Here's a subtler problem: GPUs work in 32 bits. mlx has support for float64
data types, but you can only do operations on the CPU with them. So, to work
on the GPU, we are in float32 land. That limits the precision of our results:
 
```{r}
library(RmlxStats)
base_fit <- lm(mpg ~ gear, mtcars)
mlxs_fit <- mlxs_lm(mpg ~ gear, mtcars)
max(abs(coef(mlxs_fit) - coef(base_fit)))
```
 
Does that matter? After all, differences at the 6th decimal place will often
be drowned out by sampling error.
 
```{r}


params <- expand.grid(n = 10^(5:7), r_sq = c(0.5, 0.9, 0.99))

results <- params |> 
  split(seq_len(nrow(params))) |> 
  lapply(function (param) {
    n <- param$n
    r_sq <- param$r_sq
    x <- rnorm(n)
    a <- sqrt((1 - r_sq)/r_sq)
    y <- x + a * rnorm(n)
    base_fit <- lm(y ~ x)
    mlxs_fit <- mlxs_lm(y ~ x)
    coef_diff <- max(abs(coef(base_fit) - coef(mlxs_fit)))
    se <- coef(summary(base_fit))["x", "Std. Error"]
    data.frame(n, r_sq, coef_diff, se)
  })

results <- do.call(rbind, results)

results$n <- factor(results$n)
results$`R squared` <- factor(results$r_sq)

library(ggplot2)

ggplot(results, 
       aes(coef_diff, se, colour = n, group = n)) + 
  geom_point(aes(shape = `R squared`)) + 
  geom_line() + 
  geom_function(aes(group = 1), fun = \(x) 100*x, linetype="dotted", colour = "black") + 
  scale_x_log10() + 
  scale_y_log10() +
  theme_minimal() +
  labs(
    title = "Computational error versus sampling error for mlxs_lm()",
    subtitle = "Dotted line marks where computational error is 1% of a s.e.",
    x = "Computational difference",
    y = "Standard error"
  )
```
 
 The graph above plots sampling standard errors against computational
 differences between `stats::lm()` and `RmlxStats::mlxs_lm()`. (I'm taking
 `stats::lm()` as ground truth.) Until your N
 is 10 million and your R-squared is high, sampling error dominates by far.
 
 Fine, but on the other hand...
 
 * If we are treating the coefficient of y on x as a descriptive statistic, 
   then sampling error is irrelevant.
 * Big data is RmlxStats' target market! For small datasets, base R will be
   fast enough. So we should be concerned about the 10-million-row case.
  
# High precision finish

Since recent updates, Rmlx supports float64 types. So, one possibility is
to do early work in float32 on the GPU, then finish in float64 on the CPU. 
This might give us speed *and* precision.

A natural place to try this was `mlxs_glm()`. General linear models
use iterations of least squares to find their solution, so we can iterate
on the GPU, then the CPU after we hit a tolerance threshold.

Rmlx and RmlxStats are both vibe-coded. The process has taught me about
the benefits and dangers of AI coding: the AI is a competent programmer
and extremely knowledgeable about statistics, but it will also regularly
do things that are profoundly silly, and it is quite good at hiding them.
(For example, earlier versions of Rmlx were silently ignoring the user's
choice of device.) This is particularly risky when mistakes can
lead not to an error message, but to a silently wrong statistical result!
So, before working on the code itself, I wanted to build a test harness.

I coded a series of fuzz tests, working with the AI. Its specialist knowledge
helped me do better than I could have done on my own: I would have probably
just used large datasets of random numbers, whereas it knew to focus in
on "problematic" or interesting sets of variables. It also pointed me to the
[NIST statistical datasets](https://www.nist.gov/itl/sed/products-services/statistical-reference-data-sets-strd), 
which are useful small test cases where correct answers have been calculated 
by hand.

One problem with statistical fuzz tests is knowing what the right answer is!
Mostly, I took the approach of trusting the base R implementation (or
`glmnet::glmnet()`) and taking differences from it as errors. But in some
cases we can do better - for example, by calculating performance on a
defined objective function, or testing model prediction accuracy against
an "oracle" based on the true data generating process.

Once I was happy with the fuzz tests, I stored results in a csv file
populated by a GitHub action. This lets me check how results improve
over time or with new pull requests. There's an 
[Rmarkdown vignette](https://hughjonesd.github.io/RmlxStats/articles/fuzz-test-results.html)
to show the results - it's a little crude right now, but helpful to show
when something gets better.

We then turned to `mlxs_glm()`. We added an `epsilon_f64` 
[control variable](https://hughjonesd.github.io/RmlxStats/reference/mlxs_glm_control.html) which lets the user tune when to move to float64 on the GPU.

Does it work? Here I plot speed against accuracy for different values of
`epsilon` and `epsilon_f64`.

```{r}
library(RmlxStats)
set.seed(123L)

n <- 1e5
x <- matrix(rnorm(n * 10 ), n)
y <- rnorm(n) + x %*% rep(0.1, 10)
x <- as.data.frame(x)
x$y <- y
x$y_lgl <- x$y > 0
fml <- reformulate(colnames(x)[1:10], response = "y_lgl")

base_fit <- glm(fml, data = x, family = "binomial")
base_coefs <- coef(base_fit)

eps <- 10^-(6:12)
eps_f64 <- 10^-(5:7)
params <- expand.grid(eps = eps, eps_f64 = eps_f64)
params <- params[params$eps < params$eps_f64,]

results <- params |> 
  split(seq_len(nrow(params))) |> 
  lapply(function (param) {
    nreps <- 10
    time <- system.time({
      for (rep in 1:nreps) {
        fit <- mlxs_glm(fml, data = x, family = "binomial",
                        control = list(epsilon_f64 = param$eps_f64, 
                                       epsilon = param$eps))
        coefs <- coef(fit)
      }
    })
    elapsed <- time[["elapsed"]]/nreps
    coef_diff <- max(abs(coefs - base_coefs))
    data.frame(epsilon = param$eps, epsilon_f64 = param$eps_f64,
               elapsed = elapsed, coef_diff = coef_diff)
  })

results <- do.call(rbind, results)
```

```{r}
library(forcats)

results |> 
  transform(epsilon_f64 = factor(epsilon_f64), epsilon = factor(epsilon)) |> 
  transform(epsilon = fct_rev(epsilon)) |> 
  ggplot(aes(epsilon, coef_diff, 
             size = elapsed, colour = epsilon_f64)) + 
    geom_point() + 
    scale_color_viridis_d(end = 0.9, option = "A") +
    scale_y_log10() +
    theme_minimal() + 
    labs(
      title = "Coefficient error in mlxs_glm()",
      y = "Coefficient error",
      size = "Time elapsed"
    )
```

 There's a clear jump in accuracy at 
`epsilon = 1e-7`, but this only happens when `epsilon_f64` is low
enough. Otherwise, conversion to float64 never kicks in because we simply
never hit the threshold. We never get closer than 3e-9 to the base result,
even at very low epsilons. This may simply be due to the different algorithms 
being used.

You can also see that time hasn't taken too huge a hit from increasing the
accuracy. The next graph shows this in more detail:

```{r}
library(santoku)

results |> 
  transform(
    epsilon_f64 = factor(epsilon_f64), 
    epsilon = factor(epsilon),
    Error = santoku::chop(coef_diff, 10^-(9:5), 
                             labels = lbl_glue(label = "< {r}"))) |> 
  transform(epsilon = fct_rev(epsilon)) |> 
  ggplot(aes(epsilon, elapsed, colour = epsilon_f64, group = epsilon_f64)) + 
    geom_line() + 
    geom_point(aes(shape = Error), fill = "white") + 
    scale_y_log10() + 
    scale_shape_manual(values = c("< 1e-08" = "circle", 
                                  "< 1e-06" = "circle filled")) +  
    
    theme_minimal() + 
    labs(
      title = "Speed in mlxs_glm() with float64 operations",
      y = "Time elapsed (log scale)"
    )

```

Moving to float64 too early slows you, though not by much; moving to it 
never (1e-7) is faster, but your results stay at float32 levels of accuracy.
The sweet spot is `epsilon_f64 = 1e-6`, and that is the current default. (These
are quite rough measurements for a blog post - the [benchmark](https://hughjonesd.github.io/RmlxStats/articles/fuzz-test-results.html)and 
[fuzz test](https://hughjonesd.github.io/RmlxStats/articles/benchmarks.html) 
vignettes have more details.)

# What could go wrong?

It's worth sharing a vibe-coding gone wrong anecdote. At some point, Codex asked me
if we should check for rank-deficiency in linear fits. I thought about it and said
yes, and it implemented a function `.mlxs_check_full_rank(x, ...)`, which I accepted 
without checking too hard.

Oops. The function *recomputed the QR decomposition for the design matrix `x`*, meaning we 
were now doing the most computationally expensive part of linear regression, twice.
I didn't notice the slowdown in all my benchmarks until later, and an earlier version
of this article said "`mlxs_lm()` is still slower than `stats::lm()`". No it's not!

Fortunately, Codex also saved me when I asked it to find the source of the slowdown.
It bisected commits and found the problem, leading me to fix it, and then to ask:

> You have genius-level statistical and computing knowledge. 
> Is there some instruction you could be given, that would stop you doing 
> *very stupid things*?


# Failures

I was encouraged by this, and tried similar tricks for `mlxs_lm()` and the
other package functions. None of these really worked; either
the limits on accuracy were not just down to the GPU, or my approach was wrong.
So for now, most of RmlxStats remains in float32. 

Still, the testing infrastructure I've set up should help check any future
work, and progress in the mlx framework should open up new speedups. I've also
got more comfortable working with llms. It's about finding the balance 
between speed and accuracy - kind of like the statistics, in fact.


