---
title: "Everything I know about R subsetting"
author: "David Hugh-Jones"
date: "02/08/2018"
output: 
  html_document:
    theme: readable  
    toc: true
    toc_float: true     
editor_options: 
  chunk_output_type: inline
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = '')
knitr::opts_template$set(gotcha = list(
  error = TRUE,
  class.source = "gotcha"
))
options(huxtable.knit_print_df = FALSE)
library(huxtable)
library(tibble)
```

<style type="text/css">
.gotcha {
  background: white;
  border: 2px solid red;
}

</style>


R subsetting is complex but very powerful. Or, it's powerful but very complex -- so complex
that some of what I am about to tell you, I learned only this year. (I've been using
R since 2003.) This page has everything I know about it, from the basics up. If you're an
expert, there still might be something for you here: you might want to skip to the "Advanced" 
section.

# The basics


Let's give ourselves some variables to play with -- a simple character vector with some names.

```{r}
food <- c(
        ham      = "Ham", 
        eggs     = "Eggs", 
        tomatoes = "Tomatoes"
      )
```


## Numeric, character and logical subsets


R knows three basic way to subset.

The first is the easiest: subsetting with a number `n` gives you the nth element.

```{r}
food[1]
```

If you have a vector of numbers, you get a vector of elements.

```{r}
food[2:3]
food[c(1, 3)]
```

The second is also pretty easy: if you subset with a character vector, you get the element(s) with
the corresponding name(s).

```{r}
food["ham"]
food[c("ham", "eggs")]
```

You can repeat yourself:

```{r}
cat(food[c("ham", "ham", "ham")], "! I love ham!")
```

Lastly, you can subset with a logical vector. This is a bit different: it returns every element
where the logical vector is `TRUE`.

```{r}
allergies <- c(FALSE, TRUE, FALSE)
vegetarian <- c(FALSE, TRUE, TRUE)
vegan <- c(FALSE, FALSE, TRUE)

# Vegetarian food:
food[vegetarian]
# Vegan food:
food[vegan]
```

Logical subsetting is useful when you want to find things with a particular characteristic:

```{r}
# Long food names:
nchar(food) > 5
food[nchar(food) > 5]
```

## The other three basic kinds of subsetting

Those are the three basic kinds of subsetting, depending on the type of object we use as an index.

But there are also three basic kinds of subsetting on a *different* dimension -- the syntax of the
command. So far we've just seen one syntax, using `[`.

### Subsetting with `[[`

If you want to select a single element, you can subset with double brackets: `[[]]`. 

```{r}
food[[1]]
food[["ham"]]
```

There's a subtle difference between `[` and `[[`, can you spot it?

```{r}
food[1]
food[[1]]
```

Subsetting with `[` keeps the names, subsetting with `[[` doesn't. Actually, it's a bit more than
that. Subsetting with `[` really is taking a *subset*. A subset of a named vector is a smaller
named vector. Subsetting with `[[` is taking a single *element*. Mathematicians can think of it like this:
`[` is to &sub; as `[[` is to &isin;.

This gets important when you deal with lists.

```{r}
details <- list(
        name     = "Pete", 
        birthday = as.Date("1993-01-01"), 
        spouse   = "Miranda"
      )

```

Here, `details[["spouse"]]` will give you the name `"Miranda"`. 

```{r}
wife <- details[["spouse"]]
cat("Pete's wife is", wife, ".\n")
```

But `details["spouse"]` will give you a list containing one element, named `spouse`. That
may not be what you want, because not all R commands accept lists:

```{r, opts.label = "gotcha"}
list_of_wife <- details["spouse"]
cat("This won't work:", list_of_wife)
```

Instead, you can select `list_of_wife[[1]]` to get Pete's first (and only) wife.

### Subsetting with `$`

The third kind of subsetting is a close relative of `[[`. It uses `$`. 

For example, `details$spouse` is the same as `details[["spouse"]]`. 

```{r}
details$spouse
```

The two big differences with `[[` are:

1. `$` only takes literal names. So, this works:

```{r}
date <- "birthday"
details[[date]]
```

But this doesn't, because there's no element literally called `date`:

```{r, opts.label = "gotcha"}
details$date
```

2. `$` **only** works with lists. It doesn't work with atomic vectors (vectors of numbers, characters, etc.). 

So, this won't work:

```{r, opts.label = "gotcha"}
food$ham
```

`$` is basically a useful shorthand for `[[`, when you can select an element by name.


## Recycling

So far, so easy, but there are already questions. What happens if a logical vector is shorter or
longer than what it is indexing?

The answer is that R *recycles* the vector (i.e. repeats it) along the length of the array. So, for example, `food[TRUE]` becomes `food[c(TRUE, TRUE, TRUE)]`, which gives you the whole vector
back.

```{r}
food[TRUE] 
```

If the recycling doesn't quite fit, you get a warning:

```{r}
food[c(TRUE, TRUE)] # 2 into 3 doesn't go
```

Oh! No you don't. I thought you did.

OK, but if your vector is too long, you get a warning:

```{r}
food[c(TRUE, FALSE, TRUE, TRUE)]
```

Damn.

(Goes to `?Extract`.)

(Nothing there.)

(Goes to R language definition.)

Ah, OK. If the index is longer than the vector being indexed, then the vector is "conceptually 
extended with NAs". So, above, `c(TRUE, FALSE, TRUE, TRUE)` gave us the first element,
not the second element, the third element... and an imaginary "fourth element" which is assumed to be `NA`.

## The sound of no hands clapping

While we're on the topic, what happens if you have a `NA` in your index?

```{r}
food[c(TRUE, NA, TRUE)]
food[c(1, NA, 3)]
food[c("ham", NA, "eggs")]
```

You get a `NA` in the answer. 

And if you have a zero-length index, or `NULL`, you get zero elements in the response:

```{r}
food[numeric(0)]
food[character(0)]
food[logical(0)]
food[NULL]
```

But, what if you have literally **nothing** in the index?

```{r}
food[]
```

You get everything back. **Wat.**

We're almost over the basics, and we've just got one more thing to know:

## Two dimensional subsetting

Some objects in R have two dimensions, like matrices and data frames. So you need two 
indices to take subsets of them.

Here's the food on sale at the local shop:

```{r}
(food_data <- data.frame(
  name       = food,
  price      = c(2.00, 1.50, 0.80),
  vegan      = c(FALSE, FALSE, TRUE),
  vegetarian = c(FALSE, TRUE, TRUE),
  stringsAsFactors = FALSE,
  row.names = NULL
))
```

To get the first two rows and the first four columns of this data frame, we can do:

```{r}
food_data[1:2, 1:4]
```

The first index gives the rows. The second index gives the columns. 

As we've included all the columns, there's a shorthand we could use, which is to leave out
one index altogether. 

```{r}
food_data[1:2, ]
```

Aha! Now you see why it made sense to have "nothing" select everything.

All the kinds of subset we used before still work here. For example, we can get rows 1 and 3
like this:

```{r}
food_data[c(TRUE, FALSE, TRUE), ]
```

Or we could get columns like this, by name:

```{r}
food_data[, c("name", "vegan")]
```

A typical pattern is to get columns by name and rows using a logical vector. For example, here's
how to get the names and prices of vegetarian food:

```{r}
food_data[food_data$vegetarian, c("name", "price")]
```

That seems clear enough... only, what is `food_data$vegetarian`? It must be giving
the logical index that selects only vegetarian food --- but didn't I say that `$` only
worked for lists?

I did and it does. The reason this works is that a data frame is secretly a list. It's just a list of columns. So, `food_data$vegetarian` selects one element --- i.e. one column. Here's some more examples:
```{r}
food_data$name # The 'name' column is a character vector
food_data$price # The 'price' column is a numeric vector
```

And here's how to get all foods with a price of over 1.40:

```{r}
food_data[food_data$price > 1.40, ]
```

As you might expect, the `[[` subsetting works just the same way as `$`. You can use it to 
select a single element of the data frame -- i.e. a single column. Unlike `$`, you can
use it with computed names as well as literal names:

```{r}
my_diet_preference <- "vegan"
ok_to_eat <- food_data[[my_diet_preference]]
food_data[ok_to_eat, ]
```

We could have written those last two lines as one line:

```{r, eval = FALSE}
food_data[ food_data[[my_diet_preference]], ]
```

but all those `[` characters together can make your eyes start to cross.

You can also select a single column using `[`. This is useful when you also want to select
a subset of rows:

```{r}
# Names of vegan food:
food_data[food_data$vegetarian, "name"]
```



# Advanced subsetting

Now, for some stuff you might not know.

## Don't drop that!

If a data frame is a list, and `[` selects subsets of lists, then you might have expected
`food_data[, "name"]` to select a subset of that data frame -- i.e. a smaller data frame 
with one column. That would have been consistent, and it might have saved a lot of pain
for R developers over the years... but as you can see above, it didn't happen. When you 
selected `food_data[, "name"]`, you just got a vector, not a data frame with one column.

To repeat: if you select a single column of a data frame using
`[`, you get just a vector of data. If you select two or more columns, you get the whole data frame.

This is a powerful foot shotgun.

Maybe we write a complex computation to select certain columns:

```{r}

frobnitz_the_zacular <- function () 2:3

calculated_columns <- frobnitz_the_zacular()
food_to_buy <- food_data[, calculated_columns]
nrow(food_to_buy)
```

Later on, we look through our complex code and realise we can reverse the Hyperlight Drive:

```{r}
frobnitz_the_zacular <- function () 2
```

Oh dear.

```{r, opts.label = 'gotcha'}
calculated_columns <- frobnitz_the_zacular()
food_to_buy <- food_data[, calculated_columns]
nrow(food_to_buy)
```

The way to avoid this is the special argument `drop = FALSE`. You can put this
after your rows and columns in the subset:


```{r}
calculated_columns <- frobnitz_the_zacular()
food_to_buy <- food_data[, calculated_columns, drop = FALSE]
nrow(food_to_buy)
```

You will forget this. And then, you will regret it.


## Minus numbers

Just as you use positive integers to *select* elements, 
you can use negative numbers in indices to *remove* elements.

```{r}
food[-2]
```

```{r}
food[c(-1, -3)]
```

Of course, this works for data frames too. (I won't keep saying that from now on.)

```{r}
# Everything but the first column:
food_data[, -1]
```

If you mix positive and negative numbers, you get an error. No, really.

```{r, opts.label = "gotcha"}
food[c(-2, 2)]
```

What about 0? It turns out that 0 just gets ignored.

```{r}
food[c(0, 1)]
```

## Matrix indexing

So far, we have only picked out a "square" of data from a data frame. That is, we can subset
rows and columns. But what if we want to have, say, the element `[1, 2]`, the element `[3, 3]` 
and the element `[3, 4]`?

You can do this by indexing with a 2-column matrix. The first column gives the row numbers to pick
from the original data, the second column gives the column numbers. Results will be coerced into a 
vector:

```{r}
elements <- matrix(
        c(1, 2,
          3, 3, 
          3, 4),
        nrow  = 3,
        byrow = TRUE
      )

food_data[elements]

```


## However many dimensions you want

I've assumed that one-dimensional objects get one dimensional indexing, and
two-dimensional objects get two dimensional indexing. But really, all objects
in R are one dimensional. They just pretend to have more dimensions if you
ask them nicely.

If you want to, you can just use one-dimensional indexing. We saw this already
with `food_data$vegan` and `food_data[["vegan"]]`. This treated our data frame as
a one dimensional list of columns. You can do this with `[` as well:

```{r}
food_data[1:2]
```

This does just the same thing as `food_data[, 1:2]`. On the downside, it's less clear that you are selecting columns not rows. 

On the very very up side, the following are *not* equivalent.

```{r}
food_data[, "name"] # a vector
food_data["name"] # a one-column data frame!
```

In other words, `food_data["name"]` *doesn't* drop to a vector. As it is also
easier to read than `food_data[, "name", drop = FALSE]`, this is a useful trick
for defensive code.

But there's more.

## Flattening out

Not everything with 2 dimensions is a data frame. There are also matrices. We can use 1 dimensional indexing on these too.

Here's the `euro.cross` data of conversion rates between various European currencies when the € was introduced:

```{r}
euro.cross <- round(euro.cross, 2)
euro.cross[1:3, 1:3]
rownames(euro.cross)
colnames(euro.cross)

```

We know how to pick out denmark's currency rates -- just do `euro.cross["DEM", ]`.  What if we want to pick out the currencies that are within 80% of each other's value?

```{r}
euro.cross[euro.cross < 10/8 & euro.cross > 8/10]
```

Here is how this works:

1. The index is a logical matrix.
2. Subsetting treats this as a logical vector.
3. That returns a vector of elements of `euro.cross`.
4. As the subsetting is one dimensional, the dimensions of `euro.cross` are ignored, and
the result is just a one-dimensional vector.

This may not seem so useful -- how do we know which elements we've got? But it is good if we want to
change some elements. For example, we might want to run a counterfactual and look at Eurozone
currency rates if these currencies had entered the Euro at parity with each other. Then our
data would be:

```{r}
euro.cross[euro.cross < 10/8 & euro.cross > 8/10] <- 1
euro.cross
```


How does it mean to treat a 2D object as a 1D object? That depends. 

A data frame is just a list of columns, so indexing in one dimension just gives you columns. For example, `food_data[1]` is the first column of food data.

A matrix isn't a list of columns; it's a vector with a `dim()` attribute. The vector
indexing goes down the columns. Here's an illustration:

```{r}
rc <- paste(rep(1:3, 3), rep(1:3, each = 3), sep = ",")
m <- matrix(rc, 3, 3)
```

```{r, include = FALSE}
ht <- as_huxtable(m)
ht[3, 3] <- '...'
background_color(ht)[3] <- 'blue'
background_color(ht)[5] <- 'yellow'
ht <- set_all_borders(ht, 2)
align(ht) <- 'center'
font_size(ht) <- 20
col_width(ht) <- '90pt'
width(ht) <- '270pt'
ht2 <- rbind(ht[1:3, 1], ht[1:3, 2], ht[1:3, 3])
ht2[] <- paste0(1:9, " (", rc, ")")
ht2[9, ] <- '...'

col_width(ht2) <- '90pt'
width(ht2) <- '90pt'
caption(ht) <- 'A 3x3 matrix'
caption(ht2) <- 'The matrix when indexed by a single vector'
```



```{r}
ht
```

And here's how it gets unwound by matrix indexing:

```{r}
ht2
```

So, for example, element 5 is the yellow one -- row 2 column 2.

The ability to ignore dimensions is great when you are using functions that don't
preserve dimensions themselves. For example, suppose I want to change the text of this matrix by adding some labels:

```{r}
paste("Row, column: ", m)
```

Unfortunately, `paste` doesn't preserve dimensions, so if I did 

```{r, opts.label='gotcha', eval = FALSE}
m <- paste("Row, column:", m)
```

I'd have a vector, not the matrix I wanted.

But the result is what I need. It just needs its dimensions back. I could either do `dim(m) <- c(3, 3)`, or, to avoid any errors with the dimensions wrong, I can put my
data back into the matrix shape with the "empty subset" trick:

```{r}
m[] <- paste("Row, column:", m)
m[]
```

Using `m[] <-` assigns into every element of `m`. That is different from `m<-`, which
overwrites the old object with a whole new one.

In fact, you can mix 2-dimensional subsetting with one dimensional subsetting, at your
convenience. For example, suppose I only want the "row, column" labels at the top
of every column:

```{r}
m <- matrix(rc, 3, 3)
# In 1D space:
with_labels <- paste("Row, column:", m)
# In 2D space:
dim(with_labels) <- dim(m)
m[1, ] <- with_labels[1, ]
m
```

I used this trick a lot recently, when I rewrote the [huxtable](https://hughjonesd.github.io/huxtable) package, which produces LaTeX and HTML
tables. (Huxtable produced the pictures above.) Obviously table data comes in 2D form,
but sometimes you want to work on a subset of it, treating the subset as just a vector,
then just assign back into the matrix. The pattern is:

```r
vec_subset <- matrix_data[rows, cols]
# if you need to explicitly remove the dimensions, you can do:
vec_subset <- c(vec_subset)          

# Now do a bunch of vector operations:
vec_subset <- manipulate(vec_subset) 

# Assign back into the matrix
matrix_data[rows, cols] <- vec_subset
```

Vector operations are fast, so this can be a nice way to speed up code.


## Double brackets, many dimensions

You normally only see `[[` used with one index, to select from a list. But you can use it with
multiple dimensions just like `[`. It always only returns one element:

```{r}
food_data[[1, "name"]]
```

So, if you know your code should only want one element, this is a safe way of asking for it. 


## Going deep

I never even knew about this until this year, and have never used it.

Lists are good for recursive data structures:

```{r}

details <- list(
      name = "Henry", 
      birthday = as.Date("1491-07-28"), 
      wives = list(
          Catherine = list(languages = c("Spanish", "English", children = "Mary")),
          Anne = list(motto = "Grogne qui grogne...", children = "Elizabeth"),
          Jane = list(children = "Edward"),
          Ann  = list(from = "Flanders"),
          Katherine = list(bad_habits = "flirtatious"),
          Catherine = list(religion = "Protestant") 
        )
      )
```

What if we want to dive into those structures? Say, finding the name of Henry's second wife's child?

You can deep dive into a recursive (list-like) object with `[[` and a vector index:

```{r}
details[[c("wives", "Anne", "children")]]
```

Notice that this only works with recursive objects. If you subset an atomic vector using `[[` and a vector index, you'll get an error.

# Summary

Here's a summary of how indexing works:

```{r, echo = FALSE}


index_ht <- tribble_hux(
  ~ Syntax, ~ numeric, ~ character,  ~ logical,
  "<code>[</code>",      "Subset selected by position", "Subset selected by names", 
                                     "Subset where index is <code>TRUE</code>",
  "<code>[[</code>",  "Single element selected by position", "Single element selected by name", "--",
  add_colnames = TRUE
)

index_ht <- hux(
        Vector = c("Atomic", "", "", "Recursive", "", ""),
        Index  =  rep(c("numeric", "character", "logical"), 2),
        `<code>[</code>`    =  rep(c(
                    "Subset selected by position", 
                    "Subset selected by names", 
                    "Subset where index is <code>TRUE</code>"
                  ), 2),
        `<code>[[</code>`   =  c(
                    "Single element selected by position", 
                    "Single element selected by name", 
                    "--",
                    "Single element selected by position",
                    "Single element by name (use vector for deep indexing)",
                    "--"
                  ),
        add_colnames = TRUE
      )
      
rowspan(index_ht)[c(2, 5), 1] <- 3
align(index_ht)[1, ] <- "center"
escape_contents(index_ht) <- FALSE
# right_border(index_ht)[, 1:2] <- 1
bottom_border(index_ht)[c(1, 4), ] <- 1
bottom_border(index_ht)[2, 1] <- 1
wrap(index_ht) <- TRUE
width(index_ht) <- 0.75
bold(index_ht)[1, ] <- TRUE
index_ht
```

That's all! Or at least, it's all I know. Maybe in 15 years, I'll discover some more tricks.
Meanwhile, good luck, and don't behead yourself.
